Subsidiary Recursion in Coq

### 2022-01-12 paper/paper.tex
writing up SpanAlg

Definition fold : FoldT Alg Subrec.
\end{verbatim}
\caption{The type for algebras (\texttt{Subrec.v})}
\label{fig:algf}
\end{figure}

@@

The final definition in the figure is for \verb|fold|, which allows
us to fold an \verb|Alg| over a value of type \verb|Subrec|.  We will
return to the code for this, and definitions of \verb|Subrec| and \verb|inn|,
in Section~\ref{sec:deriv}.

@@

Having seen now the interface for subsidiary recursion in Coq, let us
consider now some examples, listed by their filename in the development.

\subsection{\texttt{Span.v}}

>>>>

Definition fold : FoldT Alg Subrec.
Definition rollAlg : forall {C : Set} {X : Set -> Set}, AlgF Alg C X -> Alg C X.
Definition unrollAlg : forall {C : Set} {X : Set -> Set}, Alg C X -> AlgF Alg C X.
\end{verbatim}
\caption{The type for algebras (\texttt{Subrec.v})}
\label{fig:algf}
\end{figure}

@@

The final definitions in the figure are for \verb|fold|, which allows
us to fold an \verb|Alg| over a value of type \verb|Subrec|; and for mapping
between \verb|Alg| and its unfolding in terms of \verb|AlgF|.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

@@

Having seen the interface for subsidiary recursion in Coq, let us
consider now some examples.

\subsection{The \texttt{span} function (\texttt{Span.v})}

Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given an anchor type \verb|R : Set| along with a
fold function for that anchor type (i.e., of type \verb|FoldT (Alg (ListF A)) R|),
we would like to map an input list of type \verb|R| to a pair of
type \verb|list A * R|.  The first component of this pair is
going to be built up from scratch, and so cannot have type \verb|R|.
But the second component will be a subdatum of the input list,
and so can still have type \verb|R|.  This will enable outer recursions
to continue on that component.  So we want: 
\begin{verbatim}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{verbatim}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{verbatim}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{verbatim}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|F R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R|:
\begin{verbatim}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{verbatim}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This tail is available to the algebra
in the subdata structure of type \verb|F R|.

Figure~\ref{fig:span} gives the algebra \verb|SpanAlg| for computing
\verb|span|, and the code for \verb|spanr|.  We elide the proof
\verb|SpanFunctor| that \verb|SpanF| is indeed a \verb|Functor|,
and make \verb|X| implicit in the constructors of \verb|SpanF|.
The type of \verb|SpanAlg| is
\begin{verbatim}
Alg (ListF A) C SpanF
\end{verbatim}
This states that we are defining an algebra (\verb|Alg|)
for the \verb|ListF A| functor, with anchor type \verb|C| and carrier
\verb|SpanF|.  The algebra does not depend on what the anchor type is,
which is good, as it means we can use \verb|SpanAlg| for top-level or
subsidiary recursions.

Let us continue through the code for \verb|SpanAlg|
(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
from something whose type is an application of \verb|AlgF|.  This
takes in all the components of the recursion universe: the abstract
type \verb|R|, the \verb|reveal| function (not needed in this case),
the \verb|fold| function for any subsidiary recursions (also not
needed here), a function we choose to name \verb|span| for making
recursive calls, and finally \verb|xs : ListF A R|.  The algebra
pattern-matches on this \verb|xs|.  In the cases where it is empty or
where its head (\verb|hd|) does not satisfy \verb|p|, we return
\verb|SpanNoMatch|.  This signals to the caller that we really wished
to return \verb|([],xs)|, but could not because we do not have
\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
we recurse on the tail (\verb|tl|).  This is allowed because
\verb|tl : R| and \verb|span : R -> SpanF R|.  If \verb|span tl|
returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
we have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.  We define \verb|spanhr| to invoke the fold function it is
given, on the algebra (\verb|SpanAlg|).

The final twist is now in the definition of \verb|spanr|.  We call
\verb|spanhr| on the input \verb|xs : R|.  If \verb|spanhr| returns
\verb|SpanNoMatch|, then we are supposed to return \verb|([],xs)|,
which we can do here, because we have \verb|xs : R|.  It was only
inside the algebra that we lost the information that the subdata
structure of type \verb|F R| is derived from a value of type \verb|R|.
If \verb|spanhr| returns \verb|SpanSomeMatch|, then the return value
gives us the nonempty prefix (\verb|l|) and the suffix (\verb|r|), which we then return.

\begin{figure}
\begin{verbatim}
Definition SpanAlg(p : A -> bool)(C : Set)
  : Alg (ListF A) C SpanF :=
  rollAlg (fun R reveal fo span xs => 
     match xs with
         Nil => SpanNoMatch 
       | Cons hd tl =>
          if p hd then
            match (span tl) with
              SpanNoMatch => SpanSomeMatch [hd] tl
            | SpanSomeMatch l r => SpanSomeMatch (hd::l) r
            end
          else
            SpanNoMatch 
       end).

Definition spanhr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : SpanF R :=
  fo SpanF SpanFunctor (SpanAlg p R) xs.

Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R
  := match spanhr fo p xs with
       SpanNoMatch => ([],xs)
     | SpanSomeMatch l r => (l,r)
     end.
\end{verbatim}
\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function, and some functions based on it}
\label{fig:span}
\end{figure}
